Randomized trace estimation of parameter-dependent matrices applied to spectral density approximation

### 2024-05-19 paper/main.tex
rewrite most proofs

@@

Need for spectral density approximation

Connection to trace estimation

Analysis added to Lin Lin, also works for general matrix functions

\subsection{Notation}

$\mathbb{E}^{k}$

@@

\subsection{Parameter-dependent Girard-Hutchinson estimator}
\label{subsec:hutchinson}

\begin{theorem}{test}{test}
    $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ symmetric, continuous in $t \in [a, b]$.
    \textcolor{red}{$\delta \in (0, e^{-1})$}. Then with probability $\geq 1 - \delta$
    \begin{equation}
        \int_{a}^{b} \left| \Tr(\mtx{A}(t)) - H_{n_{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t \leq \textcolor{red}{32e^{-1}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert  ~\mathrm{d}t
    \end{equation}
\end{theorem}

Consider any time-dependent $\mtx{B}(t)$ and random variable $\vct{\psi}$ standard Gaussian
\begin{equation}
    r(t) = \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi} - \Tr(\mtx{B}(t))
\end{equation}

\begin{proof}
By \cite[Proof of Lemma 3]{cortinovis-2022-randomized-trace} we know that $r(t)$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B}(t) \rVert _F^2, 2 \lVert \mtx{B}(t) \rVert _2)$

By \cite[Theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every integer $k \geq 1$
\begin{equation}
    \mathbb{E}\left[ r(t)^{2 k} \right]
    \leq k! \left( 16 \lVert \mtx{B}(t) \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B}(t) \rVert _2 \right)^{2 k}
    = k! 2^{4 k} \lVert \mtx{B}(t) \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _2^{2 k}
\end{equation}

Since $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \geq 1$ we upper bound 
\begin{equation}
    \mathbb{E}\left[ r(t)^{2 k} \right] < \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _F^{2k}
\end{equation}

Stirling approximation \cite{robbins-1955-remark-stirling} 
\begin{equation}
    (2 k)! < 2 \sqrt{\pi k}  e^{\frac{1}{24 k}} \left(\frac{2 k}{e} \right)^{2 k}
\end{equation}

With this, we get
\begin{equation}
    \left( \mathbb{E}\left[ r(t)^{2 k} \right] \right)^{\frac{1}{2k}}
    < \left( \frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}} \right)^{\frac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B}(t) \rVert _F
\end{equation}

It can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}})^{\frac{1}{2k}} < e$ (since decreasing for $k \geq 1$).
\begin{equation}
    \left( \mathbb{E}\left[ r(t)^{2 k} \right] \right)^{\frac{1}{2k}}
    < 16 k \lVert \mtx{B}(t) \rVert _F
\end{equation}

Minkowski integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities}
\begin{equation}
    \mathbb{E}\left[ \int_{a}^{b} |r(t)|^{2 k}~\mathrm{d}t  \right]^{\frac{1}{2 k}}
    \leq \int_{a}^{b} \mathbb{E}\left[ |r(t)|^{2 k}\right]^{\frac{1}{2 k}}~\mathrm{d}t
    \leq 16 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
\end{equation}

By Markov's inequality (find citation) we have with probability at least $1 - \gamma^{-2 k}$ for some $\gamma \geq 1$
\begin{equation}
    \int_{a}^{b} |r(t)|^{2 k}~\mathrm{d}t > 16 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
\end{equation}

In particular

Use the technique from \cite[Proof of Theorem 1]{cortinovis-2022-randomized-trace}
\begin{equation}
    \mtx{B}(t)
    = \begin{pmatrix}
        \mtx{A}(t) & & \\
        & \ddots & \\
        & & \mtx{A}(t)
    \end{pmatrix}
    \qquad \text{and} \qquad
    \vct{\psi} = \begin{pmatrix}
        \vct{\psi}_1 \\
        \vdots \\
        \vct{\psi}_{n_{\Psi}}
    \end{pmatrix}
\end{equation}
Where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\Psi}}$ independent standard Gaussian. Then $\lVert \mtx{B}(t) \rVert _F = \frac{1}{\sqrt{n_{\Psi}}} \lVert \mtx{A}(t) \rVert _F$ and
\begin{equation}
    r(t) = \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Tr(\mtx{A}(t))
\end{equation}

So we conclude that with probability at least $1 - \gamma^{-2 k}$

\begin{equation}
    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Tr(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 16 k \gamma \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
\end{equation}

Setting $\delta = \gamma^{-2 k}$ (i.e. $\gamma = \delta^{-\frac{1}{2k}}$) and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\frac{1}{2}})$
\begin{equation}
    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\frac{1}{2\lceil \log(\delta^{-1}) \rceil}}
    = \lceil \log(\delta^{-1}) \rceil e^{\frac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
    \leq 2 \log(\delta^{-1}) e^{\frac{1}{2}}
\end{equation}

Consequently, with probability $1 - \delta$ we have
\begin{equation}
    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Tr(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\frac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
\end{equation}

@@

\subsection{Parameter-dependent Nyström approximation}
\label{subsec:nystrom}

Pointwise eigenvalue decomposition
\begin{equation}
    \mtx{A}(t) 
    = \mtx{U}(t) \mtx{\Lambda}(t) \mtx{U}(t)^{\top} 
\end{equation}

Partition matrices
\begin{equation}
    \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+2ex+\baselineskip}
    \mtx{U}(t) = \begin{bmatrix}
        \smash{\underbrace{\mtx{U}_1(t)}_{n \times k}} & \smash{\underbrace{\mtx{U}_2(t)}_{n \times (n-k)}}
    \end{bmatrix}
    \qquad 
    \mtx{\Lambda}(t) =
    \begin{bmatrix}
        \smash{\overbrace{\mtx{\Lambda}_1(t)}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2(t)}_{(n-k) \times (n-k)}}
    \end{bmatrix}
\end{equation}
and define $\mtx{\Omega}_1(t) = \mtx{U}_1(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2(t) = \mtx{U}_2(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$ with $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$

\begin{lemma}{OSE lemma}{ose}
    $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{(n - k) \times 2k}$ standard normal i.i.d, $\mtx{A}$, $k$ \emph{even}\textcolor{red}{(really necessary?)}

    \begin{equation}
        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] \leq C \left( \lVert \mtx{A} \rVert _2 + \frac{1}{\sqrt{k}}\lVert \mtx{A} \rVert _F \right)
    \end{equation}
\end{lemma}

\begin{proof}
Using submultiplicativity and independence
\begin{equation}
    \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    \leq \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    = \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right] \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
\end{equation}

First term $\mtx{\Omega}_2' = \mtx{\Omega}_2 / \sqrt{2k}$. This matrix has the \textcolor{red}{$(c_1\sqrt{q/m}, 9^d c_2e^{-q}, d, q)$} OSE moment property
\begin{align}
    \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2' \rVert _2 \right]
    &= \left(\mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') \rVert _2 \right]\right)^{\frac{1}{2}} \notag \\
    &\leq \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') - \mtx{A}^{\top} \mtx{A} \rVert _2 + \lVert \mtx{A}^{\top} \mtx{A} \rVert _2 \right]\right)^{\frac{1}{2}} \notag \\
    &\leq \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') - \mtx{A}^{\top} \mtx{A} \rVert _2\right]\right)^{\frac{1}{2}} + \lVert \mtx{A} \rVert _2
\end{align}

By \cite{} $\mtx{\Omega}_2'$ satisfies the \textcolor{red}{$(c_1 \sqrt{\frac{q}{m}}, c_2 e^{-q}, q)$}-JL moment property for \textcolor{red}{any $q$ (in what?)} \textcolor{red}{$\geq \dots$}, hence, by \cite[Lemma 4]{cohen-2016-optimal-approximate} it satisfies the \textcolor{red}{$(2 c_1 \sqrt{\frac{q}{m}}, 9^d c_2 e^{-q}, d, q)$}-OSE moment property for any $d \in \mathbb{N}$. Therefore, we can apply an intermediary result from \cite[Proof of Theorem 1]{cohen-2016-optimal-approximate} to bound \textcolor{red}{(order of $A\Omega$ vs $\Omega A$ ?!)}
\begin{equation}
    \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \mtx{\Omega}_2')^{\top} (\mtx{A} \mtx{\Omega}_2') - \mtx{A}^{\top} \mtx{A} \rVert _2\right] \leq \varepsilon \delta^{\frac{2}{k}}\left( \lVert \mtx{A} \rVert _2^2 + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right)
\end{equation}

Inserting this bound into the above inequality, we get
\begin{align}
    \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right]
    &= \sqrt{2k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2' \rVert _2 \right] \notag \\
    &\leq \sqrt{2k} \cdot \todo{c} \left( \lVert \mtx{A} \rVert _2^2  + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right)^{\frac{1}{2}} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2 \notag \\
    &\leq \sqrt{2k} \cdot 2 (1 + \todo{c})\left( \lVert \mtx{A} \rVert _2  + \frac{\lVert \mtx{A} \rVert _F}{\sqrt{k}} \right)
\end{align}

>>>>

\todo{
\begin{itemize}
    \item Unify index notation
    \item Double-check proofs
    \item Add all references
    \item Run numerical experiments 
    \item Write introduction / abstract
\end{itemize}
}

\begin{abstract}
    Abstract
\end{abstract}

@@

Need for spectral density approximation (cite some literature)

Connection to trace estimation (maybe only in words, not in formulas)

For \emph{many} evaluations of parameter, i.e. try to decrease complexity with each evaluation

Our work: Analysis added to Lin Lin, also works for general matrix functions

\subsection{Notation}
\begin{itemize}
    \item $e$ and $\log$
    \item the $k$-th moment of a random variable is denoted with $\mathbb{E}^{k}[X] = \left(\mathbb{E}\left[ | X |^{k} \right] \right)^{\frac{1}{k}}$;
    \item in a Gaussian random vector and a Gaussian random matrix, all entries follow independently the standard normal distribution;
    \item Schatten norms
    \item $\mathbb{N}$ all positive whole numbers
\end{itemize}

@@

We consider parameter-dependent matrices of the form
\begin{equation}
    \mtx{A}(t) = \begin{bmatrix}
        a_{11}(t) & a_{12}(t) & \dots & a_{1n}(t) \\
        a_{21}(t) & a_{22}(t) & \dots & a_{2n}(t) \\
        \vdots & \vdots & \ddots & \vdots \\
        a_{n1}(t) & a_{n2}(t) & \dots & a_{nn}(t) \\
    \end{bmatrix}.
\end{equation}
where $t$ is a parameter taking values in the interval $[a,b]$.
Mostly analogous to constant matrix.

Trace is defined as
\begin{equation}
    \Trace(\mtx{A}(t)) = \sum_{i=1}^{n} a_{ii}(t).
\end{equation}
If we can only access the parameter-dependent matrix through matrix-vector products, we can estimate the trace with the \emph{quadratic trace estimator} \cite{}: We take $n_{\Psi} \in \mathbb{N}$ Gaussian random vectors $\vct{\psi}_1,\vct{\psi}_2 \dots, \vct{\psi}_{n_{\psi}} \in \mathbb{R}^{n}$ to form
\begin{equation}
    \Hutch{n_{\Psi}}(\mtx{A}(t)) = \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j.
\end{equation}
If we gather the random vectors in a matrix $\mtx{\Psi} = [\vct{\psi}_1 ~ \vct{\psi}_2 ~ \cdots ~ \vct{\psi}_{n_{\psi}}] \in \mathbb{R}^{n \times n_{\Psi}}$ we may rewrite $\Hutch{n_{\Psi}}(\mtx{A}(t)) = n_{\Psi}^{-1} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})$. Other choices for the distribution of the entries of the random vectors are possible \cite{}. However, our theoretical developments only hold in the Gaussian case.

Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ to form the \emph{Nystr\"om approximation}
\begin{equation}
    \widehat{\mtx{A}}(t) = (\mtx{A}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} (\mtx{A}(t) \mtx{\Omega})^{\top}.
\end{equation}
Then we define the trace estimator $\Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( \widehat{\mtx{A}}(t) )$.
Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may also rewrite
\begin{equation}
    \Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{A}(t)^2 \mtx{\Omega})).
\end{equation}

Finally, a combination of the two estimators yields an analogue for the \emph{Nyström++ estimator} \cite{persson-2022-improved-variants} for parameter-dependent settings \todo{this is super ugly. find better notation}
\begin{equation}
    \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) = \Nystr{n_{\Omega}}(\mtx{A}(t)) + \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)).
\end{equation}
It is understood that $\widehat{\mtx{A}}$ is computed with the same sketching matrix as the one used for the trace estimator $\Nystr{n_{\Omega}}$. We can interpret this estimator as an interpolation between the trace of the Nystr\"om approximation and the quadratic trace estimator.

\subsection{Parameter-dependent quadratic trace estimator}
\label{subsec:hutchinson}

\begin{theorem}{Parameter-dependent quadratic trace estimator}{quadratic}
    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{n_{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 16 \gamma k \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
    \end{equation}
\end{theorem}

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to transition to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query quadratic trace estimate 
    \begin{equation}
        r(t) = \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi} - \Trace(\mtx{B}(t))
    \end{equation}
    for a parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$ in a fixed $t \in [a,b]$.
    From the proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} we know that $r(t)$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B}(t) \rVert _F^2, 2 \lVert \mtx{B}(t) \rVert _2)$. Thus, by \cite[Theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r(t)^{2 k} \right]
        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B}(t) \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B}(t) \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B}(t) \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$ we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r(t)^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\frac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments of $r(t)$ for a fixed $t$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(t) \right]
        < \left( \frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}} \right)^{\frac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B}(t) \rVert _F < 16 k \lVert \mtx{B}(t) \rVert _F,
        \label{equ:quadratic-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}})^{\frac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

    Now we make the transition to the continuous. Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities} allows us to apply \refequ{equ:quadratic-trace-onequery-fixed} in the continuous setting: 
    \begin{equation}
        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(t)|~\mathrm{d}t  \right]
        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}\left[ |r(t)|^{2 k}\right]^{\frac{1}{2 k}}~\mathrm{d}t
        < 16 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
    By Markov's inequality \cite{}, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} |r(t)|~\mathrm{d}t < 16 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
        \label{equ:quadratic-trace-onequery-uniform}
    \end{equation}

    To extend the bound to the $n_{\Psi}$-query quadratic trace estimator, we use a technique from the proof of \cite[Proof of Theorem 1]{cortinovis-2022-randomized-trace}. Let
    \begin{equation}
        \mtx{B}(t)
        = \frac{1}{n_{\Psi}} \begin{pmatrix}
            \mtx{A}(t) & & \\
            & \ddots & \\
            & & \mtx{A}(t)
        \end{pmatrix}
        \qquad \text{and} \qquad
        \vct{\psi} = \begin{pmatrix}
            \vct{\psi}_1 \\
            \vdots \\
            \vct{\psi}_{n_{\Psi}}
        \end{pmatrix}
    \end{equation}
    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\Psi}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors. Then $\lVert \mtx{B}(t) \rVert _F = \frac{1}{\sqrt{n_{\Psi}}} \lVert \mtx{A}(t) \rVert _F$ and
    \begin{equation}
        r(t) = \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)).
    \end{equation}
    Hence, we conclude from \refequ{equ:quadratic-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
        < 16 k \gamma \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
    \end{equation}

    %Setting $\delta = \gamma^{-2 k}$ (i.e. $\gamma = \delta^{-\frac{1}{2k}}$) and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\frac{1}{2}})$
    %\begin{equation}
    %    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\frac{1}{2\lceil \log(\delta^{-1}) \rceil}}
    %    = \lceil \log(\delta^{-1}) \rceil e^{\frac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
    %    \leq 2 \log(\delta^{-1}) e^{\frac{1}{2}}
    %\end{equation}
    %
    %Consequently, with probability $1 - \delta$ we have
    %\begin{equation}
    %    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\frac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
    %\end{equation}

@@

\subsection{Parameter-dependent Nystr\"om approximation}
\label{subsec:nystrom}

Pointwise eigenvalue decomposition
\begin{equation}
    \mtx{A}(t) 
    = \mtx{U}(t) \mtx{\Lambda}(t) \mtx{U}(t)^{\top} 
\end{equation}
where $\mtx{\Lambda}(t) = \operatorname{diag}(\lambda_1(t), \dots, \lambda_n(t))$.

We define the matrix partitions
\begin{equation}
    \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+2ex+\baselineskip}
    \mtx{U}(t) = \begin{bmatrix}
        \smash{\underbrace{\mtx{U}_1(t)}_{n \times k}} & \smash{\underbrace{\mtx{U}_2(t)}_{n \times (n-k)}}
    \end{bmatrix}
    \qquad 
    \mtx{\Lambda}(t) =
    \begin{bmatrix}
        \smash{\overbrace{\mtx{\Lambda}_1(t)}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2(t)}_{(n-k) \times (n-k)}}
    \end{bmatrix}
\end{equation}
and let $\mtx{\Omega}_1(t) = \mtx{U}_1(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2(t) = \mtx{U}_2(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$ with $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.

\begin{lemma}{Moment tail bound}{ose}
    A symmetric matrix $\mtx{A} \in \mathbb{R}^{l \times l}$ and stochastically independent Gaussian random matrices $\mtx{\Omega}_1 \in \mathbb{R}^{k \times 2k}, \mtx{\Omega}_2 \in \mathbb{R}^{l \times 2k}$ satisfy for all $k \in \mathbb{N}$ and a constant $c > 0$
    \begin{equation}
        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right] \leq c \left( \lVert \mtx{A} \rVert _2 + \frac{\lVert \mtx{A} \rVert _F}{\sqrt{k}} \right).
    \end{equation}
\end{lemma}

\begin{proof}
    We first use the submultiplicativity of the spectral norm and the stochastic independendence of the random matrices
    \begin{equation}
        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
        \leq \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
        = \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right] \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right].
        \label{equ:OSE-moment-bound-initial}
    \end{equation}

    To bound the first term, we define $\widetilde{\mtx{\Omega}}_2 = \mtx{\Omega}_2^{\top} / \sqrt{2k}$. We may use the triangle inequality to rewrite
    \begin{align}
        &\mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right] \notag \\
        &= \sqrt{2k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{A} \widetilde{\mtx{\Omega}}_2^{\top} \rVert _2 \right] && \text{(definition of $\widetilde{\mtx{\Omega}}_2$)} \notag \\
        &= \sqrt{2k} \cdot \left(\mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top}) (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top})^{\top} \rVert _2 \right]\right)^{\frac{1}{2}} && \text{($\lVert \mtx{B} \rVert _2 = \lVert \mtx{B} \mtx{B}^{\top} \rVert _2^{\frac{1}{2}} $ )} \notag \\
        &\leq \sqrt{2k} \cdot \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top}) (\mtx{A} \widetilde{\mtx{\Omega}}_2^{\top})^{\top} - \mtx{A}^{\top} \mtx{A} \rVert _2 + \lVert \mtx{A}^{\top} \mtx{A} \rVert _2 \right]\right)^{\frac{1}{2}} && \text{(triangle inequality)} \notag \\
        &\leq \sqrt{2k} \cdot \left( \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right]\right)^{\frac{1}{2}} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2. && \text{($(a+b)^{\frac{1}{2}} \leq a^{\frac{1}{2}} + b^{\frac{1}{2}}$)}
    \end{align}
    By \cite{} $\widetilde{\mtx{\Omega}}_2 \in \mathbb{R}^{2k \times (n - k)}$ satisfies the \textcolor{red}{$(c_1 \sqrt{\frac{q}{2k}}, c_2 e^{-q}, q)$}-JL moment property for \todo{any $q$ (in what?) $\geq \dots$} and some constants $c_1, c_2 > 0$, hence, by \cite[Lemma 4]{cohen-2016-optimal-approximate} it satisfies the \textcolor{red}{$(2 c_1 \sqrt{\frac{q}{2k}}, 9^d c_2 e^{-q}, d, q)$}-OSE moment property for any $d \in \mathbb{N}$. For convenience, we choose $d=2k$ and $q=\frac{k}{2}$, such that $\widetilde{\mtx{\Omega}}_2$ has $(c_1, c_2 (3^4e^{-\frac{1}{2}})^k, 2k, \frac{k}{2})$-OSE moment property. We may then apply an intermediary result from the proof of \cite[Theorem 1]{cohen-2016-optimal-approximate} to bound
    \begin{equation}
        \mathbb{E}^{\frac{k}{2}}\left[ \lVert (\widetilde{\mtx{\Omega}}_2 \mtx{A} )^{\top} (\widetilde{\mtx{\Omega}}_2 \mtx{A}) - \mtx{A}^{\top} \mtx{A} \rVert _2\right] \leq c_1 c_2^2 3^8e^{-1} \left( \lVert \mtx{A} \rVert _2^2 + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right).
    \end{equation}
    Inserting this bound into the above inequality and generously bounding all terms with a single constant we get
    \begin{align}
        \mathbb{E}^{k}\left[ \lVert \mtx{A} \mtx{\Omega}_2 \rVert _2 \right]
        &\leq \sqrt{2k} \cdot \sqrt{c_1 e^{-1}} c_2 3^4 \left( \lVert \mtx{A} \rVert _2^2  + \frac{\lVert \mtx{A} \rVert _F^2}{k} \right)^{\frac{1}{2}} + \sqrt{2k} \cdot \lVert \mtx{A} \rVert _2 \notag \\
        &\leq 2 \sqrt{2} (1 + \sqrt{c_1 e^{-1}} c_2 3^4)\left( \sqrt{k} \cdot \lVert \mtx{A} \rVert _2  + \lVert \mtx{A} \rVert _F \right).
        \label{equ:OSE-moment-bound-first}
    \end{align}
